The Weakly Nonlinear Magnetorotational Instability

### 2016-09-02 paper/widegap_mri_paper.tex
hmri added

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection (\citei{Newell:1969wr}), and hydrodynamic Taylor Couette flow (e.g. \citei{Recktenwald:1993}). We emphasize that this is a model equation, valid only near marginality (e.g. \citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ is in general complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (see \citei{Aranson:2002} for a thorough review).

@@

When $\xi$ in Equation \ref{eq:unperturbed_matrix_eqns} is not equal to zero, the helical MRI arises. We examine a single fiducial helical MRI case, for the parameters used by \citei{Hollerbach:2005tr}, listed in Table \ref{table:parameters}. The helical MRI is an overstability, so the ansatz linear eigenvector we consider (Equation \ref{eq:ansatz}) is characterized by a complex temporal eigenvalue $\sigma$. For our fiducial parameters, the marginal mode has a frequency $\omega = 0.153$. This means that the helical MRI modes are traveling waves, drifting in the $z$ direction.

At the conclusion of the weakly nonlinear analysis, we find that the coefficients of Equation \ref{eq:gle} are $b = $, $d = $, and $c = $, i.e. complex. The marginal helical MRI is thus described by a complex GLE. This difference in character between the amplitude equations that modulate the weakly nonlinear standard and helical MRI is a consequence of the same property that makes the helical MRI an overstability. With the introduction of an azimuthal component, the background magnetic field acquires a handedness that is not present in a purely axial field.

>>>>

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection (\citei{Newell:1969wr}), and hydrodynamic Taylor Couette flow (e.g. \citei{Recktenwald:1993}). We emphasize that this is a model equation, valid only near marginality (\citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ is in general complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (see \citei{Aranson:2002} for a thorough review).

@@

When $\xi$ in Equation \ref{eq:unperturbed_matrix_eqns} is not equal to zero, the helical MRI arises. We examine a single fiducial helical MRI case, for the parameters used by \citei{Hollerbach:2005tr}, listed in Table \ref{table:parameters}. The helical MRI is an overstability, so the ansatz linear eigenvector we consider (Equation \ref{eq:ansatz}) is characterized by a complex temporal eigenvalue $\sigma$. For our fiducial parameters, the marginal mode has a frequency $\omega = 0.153$. This means that the helical MRI modes are traveling waves, drifting in the $z$ direction with a phase velocity $\omega/k_c$.

At the conclusion of the weakly nonlinear analysis, we find that the coefficients of Equation \ref{eq:gle} are $b = $, $d = $, and $c = $, i.e. complex. The marginal helical MRI is thus described by a complex GLE. This difference in character between the amplitude equations that modulate the weakly nonlinear standard and helical MRI is a consequence of the same property that makes the helical MRI an overstability. With the introduction of an azimuthal component, the background magnetic field acquires a handedness that is not present in a purely axial field. The helical MRI eigenvectors are therefore free to be out of phase with one another. In our perturbation series, the helical MRI modes interact within and between orders with modes which carry different phases, leading to complex GLE coefficients.

The phase dynamics of the complex GLE are well-studied in a variety of systems, and depend on the values of the GLE coefficients. The complex GLE may be unstable to traveling wave instabilities such as the Benjamin-Feir instability, a generalization of the Ekhaus instability which acts on stationary waves, and to which the standard MRI system may be unstable. The complex GLE can also admit spatiotemporal chaos, and various classes of coherent structures (\citei{Aranson:2002}).
%Because the saturation amplitude of the weakly nonlinear helical MRI is governed by a complex GLE, the saturated state of the helical MRI may be unstable to 
Although a detailed description of the phase dynamics in the helical MRI is beyond the scope of this work, we note that such long-wavelength, long-timescale behavior may be observed in liquid metal helical MRI experiments.
